Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-09-27 example.tex
removing nextSib in favor of the smaller set of actions "move parent" and "move child n". this makes the math in the paper consistent with the agda formalization, which got this update earlier tonight.
i beleive, but should be checked, that i didn't mess up any of the line
number in the example. i think i also caught the text that talked about
nextsib, mostly by deleting the macro and seeing what broke.

someone else should review this commit and make sure that is correct.

we should also consider adding text how you can get next and prev sib as
derived actions, maybe in the section where we t

The actions on Lines 2-4 complete the type ascription. In particular, the
$\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and
the $\aMove{\dNext}$ action moves the cursor to the next hole.

\begin{figure}[t!]
\begin{center}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
1 &
\zwsel{\hhole{}} &
\aConstruct{\flam{x}} &
\text{(\ref{r:conelamhole})}
\\ 2 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} &
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 3 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
\aMove{\dNext{}} &
\text{({\ref{rule:move-nextSib-arr}})}
\\ 4 &
\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 5 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-arr-right})}
\\ 6 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-asc-right})}
\\ 7 &
\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} &
\aMove{\dChild} &
\text{(\ref{r:movefirstchild})}
\\ 8 &
\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
\aMove{\dChild{}} &
\text{(\ref{r:movefirstchild-lam})}
\\ 9 &
\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fvar{x}} &
\text{(\ref{r:conevar})}
\\ 10 &
\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
{\aConstruct{\fplus}}
&
\text{(\ref{rule:construct-plus-compat})}
\\ 11 &
\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fnumlit{1}} &
\text{(\ref{r:conenumnum})}
\\ 12 &
\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Constructing an increment function in Hazelnut.}
\label{fig:first-example}
\end{figure}

The actions on Lines 5-8 move the cursor to the function body. For
simplicity, Hazelnut defines only a small collection of primitive movement
actions. In practice, an editor would also define compound movement
actions, e.g. an action that moves the cursor directly to the next
remaining hole, in terms of these primitive movement actions.

Lines 9-11 complete the function body by first constructing the variable
$x$, then constructing the plus form, and finally constructing the number
$\hnum{1}$. Notice that we did not need to construct the function body in
an ``outside-in'' manner, i.e. we constructed $x$ before constructing the
outer plus form. Luckily, the transient function bodies, $x$ and $x +
\hhole{}$, could be checked against type $\tnum$ (as we will detail in
Sec. \ref{sec:holes}.)

@@

\begin{figure}[t!]
\begin{center}
\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
13 &
\zwsel{\hhole{}} &
\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}&
\text{(\ref{r:conevar})}
\\ 14 &
\zwsel{incr} &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 15 &
incr(\zwsel{\hhole{}}) &
\aConstruct{\fvar{incr}} &
\text{(\ref{r:conevar2})}
\\ 16 &
incr(\hhole{\zwsel{incr}}) &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 17 &
incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} &
\aConstruct{\fnumlit{3}} &
\text{(\ref{r:conenumnum})}
\\ 18 &
incr(\hhole{incr(\zwsel{\hnum{3}})}) &
\aMove{\dParent}&
\text{(\ref{r:moveparent-ap2})}
\\ 19 &
incr(\hhole{\zwsel{incr(\hnum{3})}}) &
\aMove{\dParent} &
\text{(\ref{r:moveparent-hole})}
\\ 20 &
incr(\zwsel{\hhole{incr(\hnum{3})}})&
\aFinish &
\text{(\ref{r:finishana})}
\\ 21 &
incr(\zwsel{incr(\hnum{3})}) &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Appyling the increment function.}
\label{fig:second-example}
\end{figure}

@@

We begin on Line 13 by constructing the variable $incr$. Line 14 then
constructs the application form with $incr$ in function position, leaving
the cursor on a hole in the argument position. Notice that we did not need
to construct the outer application form before identifying the function
being applied. Again, the transient edit state happens to be well-typed, so
we needed to make no special allowances for this order of actions.

We now need to apply $incr$ again, so we perform the same action on Line 15
as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic
structure editor, performing such an action would result in the following
edit state:
\[
incr(\zwsel{incr})
\]
This edit state is ill-typed (after \emph{cursor erasure}): the argument of
$incr$ must be of type $\tnum$ but here it is of type
$\tarr{\tnum}{\tnum}$. Hazelnut cannot allow such an edit state to arise.

The programmer could alternatively have performed the $\aConstruct{\fap}$
action on Line 15. This would result in the following edit state, which is
well-typed according to the static semantics that we will define in the
next section:
\[
incr(\hhole{}(\zwsel{\hhole{}}))
\]
The problem is that the programmer is not able to identify the intended
function before constructing the function application form. This stands in
contrast to Lines 13-14.

Hazelnut's action semantics addresses this problem: rather than disallowing
the $\aConstruct{\fvar{incr}}$ action on Line 15, it leaves $incr$ inside a
hole:
\[
incr(\hhole{\zwsel{incr}})
\]
This defers the type consistency check, exactly as an empty hole in the
same position does. One way to think about non-empty holes is as an
internalization of the ``squiggly underline'' that text or syntactic
structure editors display to indicate a type inconsistency. By
internalizing this concept, the presence of a type inconsistency does not
leave the entire program formally meaningless.

The expression inside a non-empty hole must itself be well-typed, so the
programmer can continue to edit it. Lines 16-17 proceed to apply the inner
mention of $incr$ to a number literal, $\hnum{3}$. Finally, Lines 18-19
move the cursor to the non-empty hole and Line 20 performs the $\aFinish$
action. The $\aFinish$ action removes the hole if the type of the
expression inside the hole is consistent with the expected type, as it now
is. This results in the final edit state on Line 20, as desired. In
practice, the editor might automatically perform the $\aFinish$ action as
soon as it becomes possible, but for simplicity, Hazelnut formally requires
that it be performed explicitly.

>>>>

The actions on Lines 2-5 complete the type ascription. In particular, the
$\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and
composition of the $\aMove{\dParent}$ and $\aMove{\dChildn{2}}$ actions
move the cursor to the next hole.

\begin{figure}[t!]
\begin{center}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
1 &
\zwsel{\hhole{}} &
\aConstruct{\flam{x}} &
\text{(\ref{r:conelamhole})}
\\ 2 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} &
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 3 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
\aMove{\dParent} &
\text{({\ref{rule:move-parent-arr-left}})}
\\ 4 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\hhole{}}} &
\aMove{\dChildn{2}} &
\text{({\ref{rule:move-arr-c2}})}
\\ 5 &
\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 6 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-arr-right})}
\\ 7 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
\aMove{\dParent{}} &
\text{(\ref{rule:move-parent-asc-right})}
\\ 8 &
\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} &
\aMove{\dChildn{1}} &
\text{(\ref{r:movefirstchild-asc})}
\\ 9 &
\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
\aMove{\dChildn{1}} &
\text{(\ref{r:movefirstchild-lam})}
\\ 10 &
\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fvar{x}} &
\text{(\ref{r:conevar})}
\\ 11 &
\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
{\aConstruct{\fplus}}
&
\text{(\ref{rule:construct-plus-compat})}
\\ 12 &
\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
\aConstruct{\fnumlit{1}} &
\text{(\ref{r:conenumnum})}
\\ 13 &
\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Constructing an increment function in Hazelnut.}
\label{fig:first-example}
\end{figure}

The actions on Lines 6-9 move the cursor to the function body. For
simplicity, Hazelnut defines only a small collection of primitive movement
actions. In practice, an editor would also define compound movement
actions, e.g. an action that moves the cursor directly to the next
remaining hole, in terms of these primitive movement actions.

Lines 10-12 complete the function body by first constructing the variable
$x$, then constructing the plus form, and finally constructing the number
$\hnum{1}$. Notice that we did not need to construct the function body in
an ``outside-in'' manner, i.e. we constructed $x$ before constructing the
outer plus form. Luckily, the transient function bodies, $x$ and $x +
\hhole{}$, could be checked against type $\tnum$ (as we will detail in
Sec. \ref{sec:holes}.)

@@

\begin{figure}[t!]
\begin{center}
\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} &
\textbf{Next Action} & \textbf{Rule}
\\
\hline
14 &
\zwsel{\hhole{}} &
\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}&
\text{(\ref{r:conevar})}
\\ 15 &
\zwsel{incr} &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 16 &
incr(\zwsel{\hhole{}}) &
\aConstruct{\fvar{incr}} &
\text{(\ref{r:conevar2})}
\\ 17 &
incr(\hhole{\zwsel{incr}}) &
\aConstruct{\fap} &
\text{(\ref{r:coneapfn})}
\\ 18 &
incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} &
\aConstruct{\fnumlit{3}} &
\text{(\ref{r:conenumnum})}
\\ 19 &
incr(\hhole{incr(\zwsel{\hnum{3}})}) &
\aMove{\dParent}&
\text{(\ref{r:moveparent-ap2})}
\\ 20 &
incr(\hhole{\zwsel{incr(\hnum{3})}}) &
\aMove{\dParent} &
\text{(\ref{r:moveparent-hole})}
\\ 21 &
incr(\zwsel{\hhole{incr(\hnum{3})}})&
\aFinish &
\text{(\ref{r:finishana})}
\\ 22 &
incr(\zwsel{incr(\hnum{3})}) &
\textrm{---} &
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Appyling the increment function.}
\label{fig:second-example}
\end{figure}

@@

We begin on Line 14 by constructing the variable $incr$. Line 15 then
constructs the application form with $incr$ in function position, leaving
the cursor on a hole in the argument position. Notice that we did not need
to construct the outer application form before identifying the function
being applied. Again, the transient edit state happens to be well-typed, so
we needed to make no special allowances for this order of actions.

We now need to apply $incr$ again, so we perform the same action on Line 16
as we did on Line 14, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic
structure editor, performing such an action would result in the following
edit state:
\[
incr(\zwsel{incr})
\]
This edit state is ill-typed (after \emph{cursor erasure}): the argument of
$incr$ must be of type $\tnum$ but here it is of type
$\tarr{\tnum}{\tnum}$. Hazelnut cannot allow such an edit state to arise.

The programmer could alternatively have performed the $\aConstruct{\fap}$
action on Line 16. This would result in the following edit state, which is
well-typed according to the static semantics that we will define in the
next section:
\[
incr(\hhole{}(\zwsel{\hhole{}}))
\]
The problem is that the programmer is not able to identify the intended
function before constructing the function application form. This stands in
contrast to Lines 14-15.

Hazelnut's action semantics addresses this problem: rather than disallowing
the $\aConstruct{\fvar{incr}}$ action on Line 16, it leaves $incr$ inside a
hole:
\[
incr(\hhole{\zwsel{incr}})
\]
This defers the type consistency check, exactly as an empty hole in the
same position does. One way to think about non-empty holes is as an
internalization of the ``squiggly underline'' that text or syntactic
structure editors display to indicate a type inconsistency. By
internalizing this concept, the presence of a type inconsistency does not
leave the entire program formally meaningless.

The expression inside a non-empty hole must itself be well-typed, so the
programmer can continue to edit it. Lines 17-18 proceed to apply the inner
mention of $incr$ to a number literal, $\hnum{3}$. Finally, Lines 18-19
move the cursor to the non-empty hole and Line 21 performs the $\aFinish$
action. The $\aFinish$ action removes the hole if the type of the
expression inside the hole is consistent with the expected type, as it now
is. This results in the final edit state on Line 22, as desired. In
practice, the editor might automatically perform the $\aFinish$ action as
soon as it becomes possible, but for simplicity, Hazelnut formally requires
that it be performed explicitly.
